---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sfhelpers

## Purpose

`sfhelpers` is a R package including functions helping to run routines related
primarily to simple features
([`sf`](https://cran.r-project.org/web/packages/sf/index.html))
, but also to spatial objects of
other classes / packages (
[`raster`](https://cran.r-project.org/web/packages/raster/index.html), [`sp`](https://cran.r-project.org/web/packages/sp/index.html), [`stars`](https://cran.r-project.org/web/packages/stars/index.html),
[`terra`](https://cran.r-project.org/web/packages/terra/index.html)
).
`sfhelpers` is motivated by multiple requests for code or advice related to `sf`.
By publishing `sfhelpers` I hope to at least partly satisfy these demands. `sfhelpers` is still very much in the development phase; it may or may not go
any further. Suggestions are welcome.

## Installation

``` {r eval = FALSE}
# install package devtools if not yet installed
# install.packages("devtools")

# install fast from GitHub without vignettes (not recommended)
#   devtools::install_github("a-benini/sfhelpers")

# recommended: installation from GitHub including vignettes:
 pkgs_4_vignettes <- c("magrittr", "dplyr")
 # packages required to build sfhelpers's vignettes from GitHup repository
 new_pkgs_4_vignettes <- pkgs_4_vignettes[!pkgs_4_vignettes %in% installed.packages()]
 # among these packages the ones not yet installed
 if(length(new_pkgs_4_vignettes) > 0){install.packages(new_pkgs_4_vignettes)}
 # install the still missing packages
 devtools::install_github("a-benini/sfhelpers", build_vignettes = TRUE)
 # install sfhelpers from GitHup with its vignettes
```

## Usage

```{r}
library(sfhelpers)
```

## Usage Example

`sfhelpers::st_or` is a `sf` equivalent of QGIS Union. `st_or` returns a 
geometry set containing overlapping and non-overlapping parts of two input `sf`
objects (or `sfc`s). 

```{r example}
library(sf)
library(sfhelpers)

# the two included demo sf-objects have partly identically named attribute columns
st_agr(poly_1)
st_agr(poly_2)

# st_or() suffixes by default originally identically named attribute columns
st_or(x = poly_1, y = poly_2) %>% st_drop_geometry()

st_or(poly_1, poly_2) %>% plot()
```

For the rest of the functions included in `sfhelpers` see [here](https://a-benini.github.io/sfhelpers/reference/index.html).
