---
title: "Issues with `data.table::rbindlist()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: no
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{rbindlist_issues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
Sys.setenv(LANG = "en_US.UTF-8")
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Introduction

`sf::st_as_sf(data.table::rbindllist(<list_of_sf>))` is a 
[fast alternative](https://github.com/r-spatial/sf/issues/798#)
to `do.call(rbind, <list_of_sf>)` for binding `list`s of simple features objects
(`sf`) to single `sf`. However, there are some pitfalls when `rbindlist()` is
applied to `list`s of `sf`. These are explored, and some workarounds are
presented.
 
## Packages required

```{r setup}
library(dplyr)
library(data.table)
library(sf)
```

## Compare `rbindlist()` and `do.call()`

### Inherited classes and `bbox`

We create a `list` of `sf` objects:

```{r list-sf-sfc}
nc <- st_read(system.file("gpkg/nc.gpkg", package = "sf"), quiet = TRUE)

list_of_sf <- lapply(seq_len(nrow(nc)), function(x) nc[x, ])
```

Binding such a `list` of `sf` with `do.call()` to a single `sf` object works easily
(as long as the `list` is rather small): 

```{r do-call-bind-c}
nc_do.call <- do.call(rbind, list_of_sf)
```

And will return the expected object of the class `sf`:

```{r check-do-call-returns}
all.equal(nc_do.call, nc)
```

What's different when `rbindlist()` is used instead of `do.call()`?

First of all, `rbindlist()` on its own does not bind a `list` of `sf` to an
object of the class `sf`:

```{r rbindlist-drops-sf}
rbindlist(list_of_sf) %>% class()
```

Thus, subsequently the return of `rbindlist()` needs to be converted into an 
object of the the class `sf`:

```{r rbindlist-st-as-sf-pipe}
nc_rbindlist <- rbindlist(list_of_sf) %>% st_as_sf()
```

Is the return of the pipe `rbindlist() %>% st_as_sf()` equivalent to the one of
`do.call()`?

```{r check-rbindlist-st-as-sf-return}
all.equal(nc_rbindlist, nc)
```

No, not really! The pipe-output has inherited the class `data.table`:

```{r class-sf-pipe}
class(nc)
class(nc_rbindlist)
```

Will omitting the class `data.table` eliminate the difference?

```{r omit-class-data-table}
class(nc_rbindlist) <- c("sf", "data.frame")
all.equal(nc_rbindlist, nc)
```

There's also an issue with `bbox` of the geometry column:

```{r bbox-issue}
st_bbox(nc_rbindlist) == st_bbox(nc)
```

Instead of returning the correct `bbox` for the whole geometry column,
`data.table::rbindlist()` simply copies the `bbox` of the first `list`ed
`sf` object:

```{r bbox-issue-2}
st_bbox(nc_rbindlist) == st_bbox(list_of_sf[[1]])
```

A [simple trick](https://github.com/Rdatatable/data.table/issues/4681) solves
this issue inherent to `sf`-objects compiled with `data.table::rbindlist()`:

```{r bbox-issue-3, eval = FALSE}
sf_former_dt <- sf_former_dt[1:nrow(sf_former_dt), ]
# or, preferable in a programming context:
sf_former_dt <- sf_former_dt[seq_len(nrow(sf_former_dt)), ]
```

We apply the trick ...

```{r bbox-issue-4}
nc_rbindlist <- nc_rbindlist[seq_len(nrow(nc_rbindlist)), ]
```

... check if it worked:

```{r check-2}
all.equal(nc_rbindlist, nc)
```

### CRS

`list`s containing `sf`-objects with different CRS are treated quite differently
by `do.call()` and `rbindlist() %>% st_as_sf()`. To demonstrate that, we create
such a `list`:

```{r unequal-crs}
nc <- st_read(system.file("gpkg/nc.gpkg", package = "sf"), quiet = TRUE)

nc_3857 <- st_transform(nc, 3857)

list_of_sf_with_unequal_crs <- list(nc, nc_3857)
```

`do.call()` does not bind a `list` of `sf`-objects with unequal CRS and throws a
corresponding error-message:

```{r error = TRUE}
do.call(rbind, list_of_sf_with_unequal_crs)
```

Whereas the pipe `rbindlist() %>% st_as_sf` doesn't seem to be affected: 

```{r}
sf_dt <- list_of_sf_with_unequal_crs %>% rbindlist() %>% st_as_sf()
```

The returned object of the classes `sf` and `data.table` has a CRS which is the
same as the one of first `list`ed `sf`-object:

```{r}
st_crs(sf_dt)$epsg
st_crs(sf_dt) == st_crs(nc)
```

If we reverse the sequence of elements of the `list` of `sf`-objects with
unequal CRS the newly first placed `sf`-object dictates the CRS of the by 
`rbindlist() %>% st_as_sf` bound single `sf`-object:

```{r}
sf_dt <- list_of_sf_with_unequal_crs %>% rev() %>% rbindlist() %>% st_as_sf()
st_crs(sf_dt)$epsg
st_crs(sf_dt) == st_crs(nc_3857)
```

The maloperation of `rbindlist()` regarding `list`s of `sf`-objects with
different CRS can't be handled after the fact. Thus, feeding such a `list` to
`rbindlist()` must be avoided. Hence prior checking if a `list` of `sf` contains
more than 1 CRS is crucial:

```{r}
list(nc, nc_3857) %>% lapply(st_crs) %>% n_distinct()
```


### Different `geometry_type`s

A `list` with `sf`-objects having geometry columns of different `geometry_type`s:

```{r}
l_diff_geom_type <- list(nc[1:3, ], st_cast(nc[4, ], "POLYGON", warn = FALSE))

sapply(l_diff_geom_type, st_geometry_type, by_geometry = FALSE) %>% unique()
```

`do.call()` can easily stack such a `list` and the returned `geometry_type` is
`GEOMETRY`:

```{r}
sf_do.call <- do.call(rbind, l_diff_geom_type)

st_geometry_type(sf_do.call, by_geometry = FALSE)
```

But `rbindlist()` can't handle a `list` containing different `geometry_type`s:

```{r, error=TRUE}
rbindlist(l_diff_geom_type)
```

In order to make such a `list` workable for `rbindlist()`, the `geometry_type`s
have to be homogenized:

```{r}
l_homogenized_geom_type <- lapply(l_diff_geom_type, st_cast, to = "GEOMETRY", warn = FALSE)

sapply(l_homogenized_geom_type, st_geometry_type, by_geometry = FALSE) %>% unique()
```

Once the `list`ed `sf`-objects have the same `geometry_type`, `rbindlist()` can
stack them:

```{r}
sf_rbindlist <- rbindlist(l_homogenized_geom_type) %>% st_as_sf()

st_geometry_type(sf_rbindlist, by_geometry = FALSE)
```

```{r, eval = FALSE, echo = FALSE}
sf_rbindlist            <- sf_rbindlist[1:nrow(sf_rbindlist), ]
class(sf_rbindlist)     <- c("sf", "data.frame")
row.names(sf_rbindlist) <- row.names(sf_do.call)
all.equal(sf_rbindlist, sf_do.call)
```

## Conclusions

The use of `data.table::rbindlist()` as a fast alternative to `do.call()` for
binding `list`s of `sf`-objects to a single `sf`-object is justified in
situations when `do.call()` is too slow. It does, however, require appropriate
pre- and post-treatment:

1. Check if a `list` of `sf`-objects contains more than 1 distinct CRS:

```{r eval = FALSE}
list_of_sf %>% lapply(st_crs) %>% n_distinct()
```

If so, CRS-unification needs to be done before moving on.

```{r eval = FALSE}
?st_transform
?st_crs
```

2. Check if the `list`ed `sf`-objects have different `geometry_type`s:

```{r eval = FALSE}
sapply(list_of_sf, st_geometry_type, by_geometry = FALSE) %>% unique()
```

If so, homogenize them:

```{r eval = FALSE}
list_of_sf <- lapply(list_of_sf, st_cast, to = "GEOMETRY", warn = FALSE)
```

3. Bind `list` of `sf`-objects to a single `sf`-object (which is also a
`data.table`):

```{r eval = FALSE}
list_of_sf %>% rbindlist() %>% st_as_sf() -> sf_object
```

4. Fix bounding box / `bbox`:

```{r eval = FALSE}
sf_object <- sf_object[1:nrow(sf_object), ]
# or, preferable in a programming context:
sf_object <- sf_object[seq_len(nrow(sf_object)), ]
```

5. If required, omit class `data.table`:

```{r eval = FALSE}
class(sf_object) <- c("sf", "data.frame")
```

Note that those five steps are integrated in the function
`sfhelpers::st_rbindlist()` enabling a painless and fast conversion of a `list`
of `sf` objects to a single `sf` object. Moreover `st_rbindlist()` has argument
options capable of handling `list`s of `sf` objects that differ by the names and
positions of their geometry and other attribute columns.

```{r eval = FALSE}
?sfhelpers::st_rbindlist
```

## Links

* [bind_rows_sf for combining lists of sf objects to single dataframe](https://github.com/r-spatial/sf/issues/798#)

* [rbindlist gets wrong bbox when binding a list of sf data frames](https://github.com/Rdatatable/data.table/issues/4681)

* [crs handling in combine_list_of_sf](https://github.com/r-spatial/mapedit/issues/46)
